---
title: 'Child survival for mothers: mortality change and related measures'
author: "Iván Williams and Diego Alburez-Gutierrez"
date: "07 July 2020"
output:
  html_document:
    fig_caption: yes
bibliography: bib.bibtex
---

<!--
Comentarios:
- hay dos comentarios más abajo: uno sobre algo que agregaría a la introducción y otra cosa sobre una idea interesante que podríamos agregar como mini corolario (line 217 aprox).
- Puede sonar que la instroducción es un poco larga para el resto del cuerpo. Lo dejo a tu criterio porque me cuesta sacar algo de allí, aprece todo bastante relevante.
- Perdón por mi spanglish.
- Los gráficos aún requieren trabajo. Quizás sea algo que podamos mejorar post-vista de Emilio.
- Traté de dejar expuesto que la parte de heterogeneidad es solo el planteo.
- Las referencias cruzadas no funcionan en html_output pero tengo alta confianza que lo harán en pdf.
- Resolví que en funs.R se graben los objetos numéricos, y que en cada sesión del markdown se lean. Correr el script cada vez se hace knitr hace poco dinámico el trabajo (tarda mucho, hay formas de catche pero no las exploré en profundidad). Es decir, tenemos solo dos archivos de trabajo: este markdown y funs.R.
-->

```{r setup, include = FALSE}
library(tidyverse)
library(gridExtra)
knitr::opts_chunk$set(collapse = TRUE,  echo = F, message=F, warning=F)
options(digits=2)
load("R/CS_outs.RData")
```

## Introduction

The question of child survival for mothers sits at the very center of demographic theory.
<!-- This is evidenced by the omnipresence of the 'net reproductive rate' in accounts of human and non-human populations. -->
<!-- Historical demographers draw liberally on assumptions about kin availability and individual's exposure to offspring mortality to explain fertility decisions, especially around the demographic transition [@gortfelder_survival_2020].  -->
<!-- These assumptions are often untested given the data sparsity [@livi_bacci_concise_1997;@volk_infant_2013].  -->
<!-- Economic theories of fertility that consider children as 'investment' for old-age support also rely on notions of child survival [@preston1978effects]. -->
Offspring survival is usually studied alongside fertility decline in the context of the demographic transition [@gortfelder_survival_2020], but it also matters for post-transitional societies. 
Lower fertility and higer life expectancy means that fewer children are expected to provide key emotional, social, and financial transfers to aging parents for longer periods of time.
With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
<!-- Losing an only child at old age is expected to become a more common experience in the future [@alburez_gutierrez_womens_2019]. -->
In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers [@smith-greenaway_maternal_2020].
Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.


The Covid-19 pandemic has underscored the need to understand how sudden changes in mortality affect the availability of kin. 
The elderly depend on kin support, especially during periods of lockdown and in the absence of governmental support mechanisms.
Elderly mothers may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.
How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? 
<!-- How can we characterize the timing of offspring mortality from a mother's perspective in a formal demographic framework?  -->
How do these processes play out in a heterogeneous population?



In this paper, we aim to formalize the relationship between population-level changes in mortality rates (additive or multiplicative) and changes in the lived experience of kin death.
<!-- Ivan: si decimos esto, quiza deberia ser el enfoque principale del paper tener age-specific changes?  -->
This is key for understanding the impact of age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
This paper uncovers formal relationships and provides numerical applications using data from Latin America in the second half of the twentieth century using period measures in a stable context.^[See the [Applications](#Applications) section for more details on the source, calculation procedures and main numerical results.]
<!-- A more sophisticated model using changing age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019. -->
We continue a long tradition in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations, @brass_derivation_1953 on child survival, @goodman1974 on kin survival in stable populations, and recently, @caswell_formal_2019 kinship matrices.
<!-- The 'counting method' approach was further popularized in @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'. -->


<!-- starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes.  -->
<!-- To the best of our knowledge, @brass_derivation_1953 first proposed a formula to represent child survival over age.  -->
<!-- The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.).  -->
<!-- The 'counting method' approach was further popularized in @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'. -->

<!-- In some place (maybe here) I would say something like this: "
These work can be read like an humble extension of the kin expressions developed by formal demograperhs mentioned before, motivated by the study of beaverement experience in mothers.
(como una forma de decir: miren lo dificil ya esta hecho, nosotros continuamos un camino, dando algun que otro pequeño paso motivados por...)" -->


## Relationship (and proofs)

### Child Survival

Let $CS_a$ be the expected number of surviving children to a mother^[We use woman and mother interchangeably in this paper, assuming that all women are exposed to the same fertility rates.] alive at aged $a$ in a female stable population with fertility rates $m_{x}$, mortality hazard $\mu_{x}$ and survival function $l_x=e^{-\int_{0}^{x}{\mu_t}\,dt}$ (with unit radix $l_0=1$), as proposed by @goodman1974:

$$CS_a = \int_{0}^a{m_{x}l_{a-x}dx}$$

### An approximation

As an initial step, we provide an intuitive understanding of the effect of changing mortality on child survival. 
Building on work by @KeyfitzMath for the probability of a living mother, we can show which features of the child survival function explain the expected number of children. 
For this we use an approximate of $l_x$ using Taylor´s theorem until second order around the mean age of childbearing $\kappa$:

$$CS_a \approx l_{a-\kappa} \int_{0}^a{m_{x}dx}+(l_{a-\kappa})^{'}\int_{0}^a{(x-\kappa)m_{x}dx}+
(l_{a-\kappa})^{''}\int_{0}^a{\frac{(x-\kappa)^2}{2} m_{x}dx}\\
  \approx F_a l_{a-\kappa} + \frac{\sigma^2}{2}F_a (l_{a-\kappa})^{''}\\
  \approx F_a \, l_{a-\kappa}[1+\frac{\sigma^2}{2}\frac{(l_{a-\kappa})^{''}}{l_{a-\kappa}}]$$

where the fertility pattern by age is concentrated around $\kappa$ and the accumulated fertility (or gross reproduction rate in our one-sex population) is $F_a =  \int_{\alpha}^a{m_{x}dx}$. The second Taylor´s term is null because $\int_{\alpha}^a{x \, m_{x}dx} = \kappa \, F_a$. 

We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[Different from the findings of @KeyfitzMath of having living ancestors.] and the survival from birth to $a-\kappa$. 
The approximation is affected (negativally) by the disperssion of fertility by age (variance $\sigma^2$).
The same is true for the curvature of the survival surve, in an age range where $l_x$ is tipycally very flat (range 20-40 years old) in transitioned populations^[This approximation is useful also for get a dimension changes in the fertility average age \kappa: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx - \mu_{a-\kappa} \Delta\kappa$]. 

### Absolute Mortality Change

We now consider the consequences of an absolute, or additive, change $\delta$ in mortality in the range $[0,a-\alpha]$, where $\alpha$ is the start age of fertility risk.
For now, let us consider $m_{x,\delta}=m_{x}+\delta$ (@wrycza2012) and $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{(\mu_t+\delta})dt}$:

$$CS_{a}^\delta = \int_{\alpha}^{a}m_{x} l_{a-x} e^{-\delta (a-x)} dx.$$

We get the derivative of $dCS_{(a)}^\delta / d\delta$ evaluated near zero (@KeyfitzMath) to find the effects of adding $\delta$ to the hazard rates:

$$\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} m_{x} {l_{a-x}x dx}\\
= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}$$

Dividing both sides by $CS_{a}$ we obtain a discrete approximation:

$$\frac{\Delta CS_{a}}{CS_{a}} \approx -(a - k_a) \Delta\delta.$$

The expected change in descendants survival is inversely proportional to the difference between maternal age $a$ and the mean age of the mother at the birth of her surviving daughters $k_a$. The magnitude of the change depends negatively on the age distribution of the surviving offspring. This is intuitive considering that younger descendants experiences longer periods of exposure to risk. 

However, it is unlikely for mortality to change at the same rate at all ages. 
In the course of the demographic transition, for example, we owuld expect larger changes in young-age mortality.
As a corollary, we consider an additive change in infant mortality in the age range $\left[0;1\right)$. 
For this, we split the integral to considering those born in the previous year for $a>1$:

$$CS_{a}^{\delta_0} = \int_{0}^{a-1} m_{x} {e{^{-\int_{1}^{a-x}\mu_t \,dt-\int_{0}^{1}(\mu_t + \delta)\,dt}}} dx + \int_{a-1}^{a} m_{x} e^{-\int_{0}^{a-x}(\mu_t+ \delta)\,dt}dx\\
=\int_{0}^{a-1} m_{x} l_{a-x}e^{-\delta} dx + \int_{a-1}^{a} m_{x} l_{a-x} e^{-\delta(a-x)}\,dx$$

Deriving by $\delta$ and valuating near 0, we get:

$$CS_{a}^{\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
Since $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and finally divide by $CS_a$: 
$$ = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\\
\frac{\Delta CS_{a}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a} (1-(a-\kappa_{a-1,a}))\right] \Delta\delta$$


<!-- Este parrafo necesita mas claridad -->
An absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cummulative alive daughters for the same period, weighting how important was this in total successful experience (in terms of alive descendants).

### Proportional Mortality Change

We now consider the consequences of a proportional change in mortality on child survival.
Given a proportional change in mortality $\mu_{x,\delta}=\mu_{x}(1+\delta)$ so that $l_{a-x}^\delta = e^{-\int_{0}^{a-x}{\mu_t(1+\delta})dt}=(l_{a-x})^{(1+\delta)}$:

$$CS_{a}^\delta = \int_{\alpha}^{a} {m_{x} l_{a-x}^{(1+\delta)}} dx.$$

Using the derivative $\frac{dl_{a-x}^{(1+\delta)}}{d\delta} = log(l_{a-x}) l_{a-x}^{(1+\delta)}$, and in the third row reversing integrals between $t$ and $x$:

$$\frac{d CS_{a}^\delta}{d \delta} = \int_{0}^{a} {m_{x} l_{a-x} \log(l_{a-x}) dx}\\
= - \int_{0}^{a} {m_{x} l_{a-x} \int_{0}^{a-x}{\mu_t \, dt}\, dx}\\
= - \int_{0}^{a}{\mu_t} \int_{0}^{a-t} {m_{x} l_{a-t-x} \, dx}\, dt$$

Considering that the last integral is equal to $CS_{a-t}$, we divide by $CS_a$ and multiplying by $\delta$ we obtain:  

$$\frac{\Delta CS_{a}}{CS_a} \approx - \left[\int_{0}^{a}{\mu_x  \, \frac{CS_{a-x}}{CS_a} \,dx \,}\right] \Delta \delta$$

<!-- This needs more clarity -->
This is the negative of cumulative hazard $H_a$ but considering a positive factor $\frac{CS_{a-t}}{CS_a}$ that takes in account the relative number of surviving descendants that would be lost at each child age at risk. 
Grasping this relationship intuitively is more difficult given the interaction of birth and mortality rates (@KeyfitzMath). 
This factor has a S-shape because $l_x$ curvature and fertility accumulation, that gives more weight to first ages, generally less than 1 except for mother ages close to age valuation *a* with complete reproductive life and high infant mortality (no additional children and high risk for one age additional of survive), typically pretransitional countries with a near 50. 
We provide a numerical evaluation of both analytical expressions (absolute an relative) in the [Applications](#Applications) section.   

## Related Measures

### Burden and timing of maternal bereavement


We call *Mean Time Spent in Bereavement* (*MTSB*) the absolute measure of the expected total lifetime with a death daughter for a mother aged *a*, which can be expressed in terms of a temporary expected lost years index, in line with $e^\dagger$ (@Vaupel1986):

$$MTSB_a = \int_0^a{m_x\int_0^{a-x}{d_t e_{0|a-x-t}\, dt}\,dx} = \int_\alpha^a{m_x e_{0|a-x}^\dagger \,dx}$$.

Where $d_t$ is the death distribution from birth, $e_{0|a-x-t}$ is the life expectancy at birth until age $a-x-t$ and $e_{0|a-x}^\dagger$ the temporary dispersion measure. But most interesting would be to compare these years with the time that mothers would expect to live with their daughters. We call this the *Intensity Time in Bereavement* ($ITB$): a ratio between expected time with a "lost" and expected time with a "life", that allows to make comparisons between population regimes.

$$ITL_a = \frac{\int_0^a{m_x e_{0|{a-x}}^\dagger dx}}{\int_0^a{m_x e_{0|a-x} dx}}$$

This is a ratio between child-years in two radically different states. Looks similar to the transcendental entropy measure $H$ (@KeyfitzMath) but considering all the cohorts born during the mother´s life, weighted by their relative size $m_x$. In the figure \@ref(fig:plot_ITL_MAL) is shown that *ITB* is bigger for young women because of the weight of infant mortality in their mother experience, also with more dispersion for same levels of parity. 
 
Another important factor for the child survival experience of mothers, is the mean age at child loss, called here $MAL$. This relation can be derived by starting with the mother age $x+t$ at each death child age $t$ at death, weighted by the fertility and survival function. In it, $\kappa$ is the mean age at childbirth for women aged $a$, $MAD_{a-x}$ refers to the mean age at death for newborns that die before $a-x$ and $F_a$ is the accumulated fertility for a women aged $a$:

$$MAL_a = \frac{\int_0^a{m_x {\frac{\int_0^{a-x} l_t \, \mu_t (x+t) dt}{\int_0^{a-x} l_t \, \mu_t dt}}dx}}{\int_0^a{m_x}dx} \\
  MAL_a = \frac{\int_0^a{m_x \left[{x+\frac{\int_0^{a-x} l_t \, \mu_t \, t \, dt}{\int_0^{a-x} l_t \, \mu_t dt}}\right]dx}}{\int_0^a{m_x}} \\
  MAL_a = \frac{\int_0^a{m_x \,x \,dx}}{F_a} + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a} \\
  MAL_a = \kappa + \frac{\int_0^a{m_x MAD_{a-x}dx}}{F_a}$$

A numerical approximation of both measures are done in [Applications](#Applications) section.  

### Heterogeneity (first ideas, some guidance is needed)

In a model with heterogeneity, determined at birth by a multiplicative effect, $CS_a$ could be interpreted as a conditional expectation with random variables *K* for fertility heterogeneity and *Z* for mortality frailty.

$$CS_a(k,z) = \int_{0}^{\infty}\,{m(k)}_x\,{l(z)}_{a-x}\,dx$$

Following @Coresh1988 we can express the fertility part in a multiplicative way with variability in the level but not in the shape, as ${m(k)}_x = F_\beta\,k\,r_x$, where $F_\beta$ is the baseline cumulated fertility until upper limit in all age range until $\beta$, $k$ is a random variable with mean 1 that allows variability between groups and $r_x$ is the fertility structure by age ($\int_{0}^{\beta}{r_x}=1$). 
In the other side, frailty part can be thought in a cohort effect way, also in the multiplicative assumption as @Vaupel2014, with $l(z)_x=e^{-H_{x}z}=l_x^z$, with baseline hazard $\mu$.
Considering the joint distribution $f_{kz}$, the intuition could be expressed as  $f_{zk}=f_{k|z}f_z$: groups with higher descendant´s mortality would adjust their fertility level, with positive correlation.
Replacing both variables, the unconditional mean $\overline{CS}_a$ would be^[Note that given k and z values, fertility and child survival are independent.]:

$$\overline{CS}_a = \int_{0}^{\infty}\int_{0}^{\infty}\left[\int_{0}^{a}{ {m(k)}_x l(z)_{a-x}\,dx}\right]f_{kz}\,dz \,dk\\
= F_\beta \int_{0}^{a-x} r_x \int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z} \, k\, f_{kz} \,dk\,dz\,dx$$

Isolating for an age *x* we can inspect the part related to z and k: $(l_{a-x})^z \, k\, f_{kz}$. 
Following @Gupta2018, one way to create a correlated bivariate distribution is assume for example that $Z = Y_0 + Y_1$ and $K = e^{(Y_0 + Y_2)}$, given the fact that the historical relation is not linear between both components (see figure \@ref(fig:plot_tfr_q0), a funtion *f* like $F_\beta=f(ln(q_{0,5}))$. The auxiliar variables $Y_0$, $Y_1$ and $Y_2$ are independet and Gamma distribuited. Then we can express, using Laplace transform properties:

$$\int_{0}^{\infty} \int_{0}^{\infty} {(l_{a-x})^z \, k\, f_{kz} \,dk\,dz}\\
\int_{0}^{\infty} \int_{0}^{\infty}{e^{-H_x(Y_0 + Y_1)}e^{Y_0 + Y_2}}\\
L_{Y_0}[1-H_x]L_{Y_1}[-H_x]L_{Y_2}\,dY_0\,dY_1\,dY_2$$

...

*What can we look for here?*:

* _a formalization of covariance_

* _estimate joint distribution using historical data. Idea: ¿maybe taking all the period/country observations as realized values of an hypothetical stable population, and using bayesian procedures?_

...

## Applications {#Applications}
### Data {#Data}

We provide empirical estimates using fertility and mortality rates from the 2019 Revision of the UN World Population Prospects (UN WPP) for the Latin American Region (@WPP19). We smoothed female $l_x$ in quinquennial ages, using cubic-splines constrained to monotonic decrease, taking $L_0$ and $T_{100}$ from raw life tables as inputs for year-person calculations. For splitting fertility five groups was used quadratic optimization approach by @Michalski2018, with an desirable property for our purpose which is a good fitting in parity. Also was assumed an unique female percentage of newborns of `r round(im, 4)` for all period-country cases. Calculations were done in a discrete way assuming that the $m_x$ live births are born at exact mother´s age *x*.
An already known log relation between infant mortality and TFR is shown in Latinoamerican experience (\@ref(fig:plot_tfr_q0)). 

<!-- Specific paths like Argentina and Guatemala shows very different demographic profiles during second part of XX century. -->

```{r plot_tfr_q0, fig.cap="Female probability of death at birth by gross reproduction rate. Latin America countries in period 1950-2015"}

ggplot(LA %>% filter(x==0, name!="Latin America and the Caribbean"), 
                  aes(x=qx, y=tfr))+
                  geom_point(aes(color=Year)) +
                  geom_smooth(se = F, color = "grey", method = "lm",
                              formula = y ~ log(x), linetype = "dashed") +
                  geom_line(data = LA %>% filter(x==0, name=="Latin America and the Caribbean"),
                            aes(x=qx, y=tfr, color=Year), size=1)+
                  theme_classic() + labs(x="q(0)", y="GRR") + 
                  theme(legend.position = "none") 
                  # geom_line(data = LA %>% filter(x==0, name %in% c("Guatemala")),
                  #           aes(group=name), colour=2, size = 1) +
                  # geom_line(data = LA %>% filter(x==0, name %in% c("Argentina")),
                  #           aes(group=name), colour=4, size = 1)
```


### Numerical Results

Consider a Latin American woman standing before us.
If this were 1950-1955, she could reasonable expect to have `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` surviving daughters on her 50$^{th}$ birthday.
In 2010-2015 a woman the same age would only have `r mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` living children. 
The difference of `r mean(CS_outs$CS[CS_outs$Period == "1950-1955" & CS_outs$Age==50])-mean(CS_outs$CS[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` children is explained by reduced fertility and improved mortality in the region [give stats]. 
We now remove the effect of changing fertility by considering the number of daughters surviving up to maternal age $a$ as a proportion of the daughters ever born to a woman that age, $\frac{CS_a}{F_a}$. 
Given Eq. \ref{eq:CS}, the increase from `r mean(CS_outs$CS_prob[CS_outs$Period == "1950-1955" & CS_outs$Age==50])` in 1950-1955 to `r mean(CS_outs$CS_prob[CS_outs$Period == "2010-2015" & CS_outs$Age==50])` in 2010-2015 must be explained by a change in mortality.

<!-- this next paragrap es una idea.can be easily solve with optimization. Let´s discuss what you think.
$CS_a$ is usually an increasong function of age, with a pick An interesting corollary of these results is the question: at wich ages a mother would have the same *CS*, like a mirror effect? if we consider age *a*, the question could be answered by solving numerically *y* for: $\int_{0}^{a}{m_xl_{a-x}}dx=\int_{0}^{y}{m_xl_{y-x}}dx$, or with our approximation estimating *y* that $\frac{F_a}{F_y} = \frac{l_{a-\kappa_a}}{l_{y-\kappa_y}}$.
```{r, include=F}
# Por ejemplo... Una mujer de Bolivia en 1950-1955 esperaba tener la misma cantidad de hijos vivos a los 70 que a los 37.
CS_outs %>% filter(Period=="1950-1955", name=="Bolivia") %>% select(Age,CS)
```
End idea-->

```{r plot_CS, fig.cap="Child Survival and Child survival as a share of cummulative fertility by age, for women aged 30, 40 and 50. Estimates using UN WPP data for Latin American countries in the period 1950-2015 period"}
a <- ggplot(CS_outs %>% filter(name!="Latin America and the Caribbean"), 
            aes(x=Year, y=CS, color=factor(Age)))+
            geom_point(alpha = .3) + 
            geom_line(data = CS_outs %>% filter(name=="Latin America and the Caribbean"), 
                      aes(x=Year, y=CS, color=factor(Age))) +
            theme_classic() +
            labs(y = "CS") + theme(legend.position = "none")
b <- ggplot(CS_outs %>% filter(name!="Latin America and the Caribbean"), 
            aes(x=Year, y=CS_prob, color=factor(Age)))+
            geom_point(alpha = .3) + 
            geom_line(data = CS_outs %>% filter(name=="Latin America and the Caribbean"), 
                      aes(x=Year, y=CS_prob, color=factor(Age))) +
            theme_classic() +
            labs(y = "CS/F") + 
            scale_colour_discrete(name="Age")
grid.arrange(a,b,ncol=2)
```

The approximation proposed for CS_a is reaalyprecise, especially for recent periods. For exaple, Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).

```{r plot_CS_aprox, fig.cap = "Left: Error in approximation for a=40, from years 1950 (darker) to 2015 (lighter) for all Latin American countries (blue) and Guatemala (red). Right: Change in survival and fertility by age in Guatemala."}
age = 40
CS_app <- ggplot(CS_outs %>% filter(Age == age, name!="Latin America and the Caribbean"), 
            aes(x=CS, y=error_aprox, color=Year))+
            geom_point(alpha = .6) + 
            geom_point(data = CS_outs %>% filter(name=="Guatemala",Age == age),
                       aes(x=CS, y=error_aprox), color="red") +
            geom_line(data = CS_outs %>% filter(Age == age, name=="Latin America and the Caribbean"), 
                      aes(x=CS, y=error_aprox), color="magenta1") +
            geom_hline(yintercept = 0, color="grey", linetype = 2) +
            theme_classic() +
            theme(legend.position = "none") +
            labs(y = "%")
Guate_lt <- LA %>% filter(name=="Guatemala") %>% ggplot() +
            geom_line(aes(x=x, y=lx/100000, color=Period)) +
            theme_classic() +
            scale_color_grey()+
            theme(legend.position = "none") +
            labs(y = "")
Guate_mx <- LA %>% filter(name=="Guatemala", between(x, 15,49)) %>% ggplot() +
            geom_line(aes(x=x, y=asfr/tfr*100, color=Period)) +
            theme_classic() +
            scale_color_grey()+
            theme(legend.position = "none") +
            labs(y = "%")
grid.arrange(CS_app, grid.arrange(Guate_lt,Guate_mx,ncol=1), ncol=2)
```

The figure \@ref(fig:CS_abs_app) shows that goodness of fit is decreasing with the change size, given that $\delta$ is assumed near zero. To ilustrate that considering $\delta=0.01$ means at age 50, it means for the absolute change case that $\frac{l_{50}}{e^{\delta\,50}} = l_{50}/$`r round(exp((50*0.01)),1)`, a big change). For the relative case was assumed a constant within each interval $\mu_{x+t}=\mu_x$ for $t$ between 0 and 1, $\int_{x}^{x+1}{\mu_t dt} = \log(l_{x+1})-\log(l_{x})$, and was used this approximation on the empirical data $\frac{\Delta CS_{a}}{CS_a} \approx - \left[\sum_{0}^{a-1}{[\log(l_{x+1})-\log(l_{x})]\, \frac{CS_{a-1-x}}{CS_a}\,}\right] \Delta \delta$.

```{r CS_abs_app, fig.cap= "Effect of an absolute absolute change in mortality on child survival by maternal age of women in Latin America  for the 1950-2015 period. Goodness of fit of approximation compared to direct estimation"}

ch_abs <-  CS_change_app %>%
              ggplot() +
             geom_point(aes(x=change_obs_abs, y=change_app_abs, color=delta)) +
             geom_abline(slope=1, intercept = 0) +
             scale_x_continuous("Empirical change in CS") +
             scale_y_continuous("Approximated change in CS") +
             scale_color_continuous("Delta") +
             theme_classic()

ch_rel <-  CS_change_app %>%
              ggplot() +
             geom_point(aes(x=change_obs_rel, y=change_app_rel, color=delta)) +
             geom_abline(slope=1, intercept = 0) +
             scale_x_continuous("Empirical change in CS") +
             scale_y_continuous("Approximated change in CS") +
             scale_color_continuous("Delta") +
             theme_classic()

grid.arrange(ch_abs, ch_rel, ncol=2)

  
```


Assuming constant period rates the expected time in years that a mother aged 30 passed with a death son was around 4% in some countries at middle XX Century. When increasing age, the survival experience depends less on infant mortality, and the distribution is around less than 2% for women aged 50, converging to 0 on time. As extreme cases, in 1950-1955 Haiti women aged 30 would have experienced an intensity of `r round(CS_outs$ITL[CS_outs$name=="Haiti" & CS_outs$Age==30 & CS_outs$Period=="1950-1955"],1)`%, and in 2010-2015 `r round(CS_outs$ITL[CS_outs$name=="Haiti" &CS_outs$Age==30 & CS_outs$Period=="2010-2015"],1)`%, while the women of Costa Rica `r round(CS_outs$ITL[CS_outs$name=="Costa Rica" & CS_outs$Age==30 & CS_outs$Period=="1950-1955"],1)`%, and in 2010-2015 `r round(CS_outs$ITL[CS_outs$name=="Costa Rica" &CS_outs$Age==30 & CS_outs$Period=="2010-2015"],1)`%.
For those Latin American mothers aged 50 with 3 daughters born who suffered a lost, they experienced that at age 30 in average (figure \@ref(fig:plot_ITL_MAL)).

```{r plot_ITL_MAL, fig.cap = "a) Intensity Time Lost of  women aged 30, 40 and 50. b) Mean age at Lost for women aged 40. Years 1950 (darker) to 2015 (lighter). Latinamerican countries in period 1950-2015"}
ITL <- ggplot(CS_outs %>% filter(name!="Latin America and the Caribbean"), 
          aes(x=Fa, y=ITL, color=factor(Age)), alpha = .3) + # alpha not working
          geom_point() +
          geom_line(data = CS_outs %>% filter(name=="Latin America and the Caribbean"), 
                    aes(x=Fa, y=ITL, color=factor(Age)), size =1)+
          theme_classic() +
          theme(legend.position = "right") +
          labs(y = "%")+
          scale_color_discrete(name = "Age")

age = 50
MAL <- ggplot(CS_outs %>% filter(Age==age, name!="Latin America and the Caribbean"))+
        geom_point(aes(x=Fa, y=MAL, color=Year), alpha = .3) +
        geom_line(data = CS_outs %>% filter(Age==age, name=="Latin America and the Caribbean"),
                  aes(x=Fa, y=MAL, color=Year),size= 1) +
        theme_classic()+
        theme(legend.position = "none") +
        labs(y = "Mean Age at Lost")
grid.arrange(ITL, MAL, ncol=2)
```


## References